---
title: "Predicting Breast Cancer Malignancy"
author: "Xinyu Xie, Jackie Yang, Xu Hao"
date: "May 9, 2025"
output: 
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    always_allow_html: true
    number_sections: false 
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5)
```


```{r message = FALSE, warning = FALSE, echo = TRUE}
rm(list = ls())
library(tidyverse)
library(kableExtra)
library(caret)
library(glmnet)
library(pROC)
library(randomForest)
library(class)
library(factoextra)
library(lme4)
Cancer_Data <- read.csv("Cancer_Data.csv")
```

## Dataset Overview

**Research Question** Can a derived tumor volume feature lead to a simpler and more interpretable generalized linear model (GLM) for predicting tumor malignancy?

**Main features**

`id`: Represents a unique ID of each patient

`diagnosis`: Indicates the type of cancer. "M" represents malignant, and "B" represents benign.

`radius_mean`: mean of distances from center to points on the perimeter

`texture_mean`: standard deviation of gray-scale values

`area_mean`: Mean area of the cell nuclei (measured in pixels)

`perimeter_mean`: mean size of core tumor

`smoothness_mean`: mean of local variation in radius lengths

`compactness_mean`: mean of $\frac{\text{Perimeter}^2}{\text{Area} - 1.0}$

`concavity_mean`: mean of severity of concave portions of the contour

`concave points_mean`: mean for number of concave portions of the contour, represents the mean values of the cancer's visual characteristics

Response Variable 'diagnosis' is a categorical variable with 2 levels ("M" for malignant and "B" for benign). All other predictor variables are quantitative, capturing various characteristics of cell nuclei.

A subset of quantitative features—`radius_mean`, `perimeter_mean`, and `area_mean`—served as primary predictors in the modeling process.



**Resources**  

- Dataset: [Kaggle – Cancer Data](https://www.kaggle.com/datasets/erdemtaha/cancer-data)  

- Reference: [Volume-Diameter Article on JCPSP](https://jcpsp.pk/article-detail/comparison-of-tumor-volume-dependent-mean-diameter-with-the-longest-diameter-and-tumor-volume-in-assessing-axillary-lymph-nodes-in-breast-cancer)


## Table Preview: Selected Predictors for Diagnosis Modeling

```{r}
Cancer_Data %>%
  select(diagnosis, radius_mean, texture_mean, area_mean, 
         smoothness_mean, compactness_mean, concavity_mean, concave.points_mean) %>%
  head(5) %>%
  kable(caption = "First 5 Rows of Key Variables Used in Modeling") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered"))
```

## Data Preparation

```{r}
#Clean and transform dataset
Cancer_Data <- Cancer_Data %>%
  dplyr::select(-id, -X) %>%
  na.omit() %>%
  mutate(
    tumor_volume = ((pi / 6) * radius_mean * 
                    (perimeter_mean / (2 * pi)) * 
                    sqrt(area_mean / pi)),
    diagnosis_bin = ifelse(diagnosis == "M", 1, 0)
  )

#Standardize selected predictors
xvars <- c(
  "radius_mean", "texture_mean", "perimeter_mean", "area_mean", "smoothness_mean",
  "compactness_mean", "concavity_mean", "concave.points_mean", "symmetry_mean", "fractal_dimension_mean",
  "radius_se", "texture_se", "perimeter_se", "area_se", "smoothness_se",
  "compactness_se", "concavity_se", "concave.points_se", "symmetry_se", "fractal_dimension_se",
  "radius_worst", "texture_worst", "perimeter_worst", "area_worst", "smoothness_worst",
  "compactness_worst", "concavity_worst", "concave.points_worst", "symmetry_worst", "fractal_dimension_worst",
  "tumor_volume"
)
Cancer_Data[, xvars] <- scale(Cancer_Data[, xvars], center = TRUE, scale = TRUE)

#Preview cleaned dataset
summary(Cancer_Data)
```


### Planned Modeling Formula

Based on our exploratory analysis, we plan to fit the following logistic regression model:

$$diagnosis_{bin} \sim concave.points_{worst} + tumor_{volume} + texture_{worst} + smoothness_{worst}$$


## Exploratory Plots for Key Predictors

The following boxplots illustrate the distribution of selected predictors stratified by tumor diagnosis. These visualizations support the relevance of each variable in distinguishing malignant from benign tumors.

### Radius Mean by Diagnosis

```{r}
#Colorpalette
diagnosis_colors <- c("B" = "#00BFC4", "M" = "#F8766D")

ggplot(Cancer_Data, aes(x = diagnosis, y = radius_mean, fill = diagnosis)) +
  geom_boxplot() +
  scale_fill_manual(values = diagnosis_colors) +
  labs(title = "Radius Mean by Diagnosis", x = "Diagnosis", y = "Radius Mean")
```


Malignant tumors show significantly higher radius_mean values (median = 0.91) than benign ones (median = -0.55). This indicates that tumor size—specifically the distance from center to edge—is a strong differentiator between classes, and radius_mean serves as a critical feature in model construction.


### Perimeter Mean by Diagnosis

```{r}
ggplot(Cancer_Data, aes(x = diagnosis, y = perimeter_mean, fill = diagnosis)) +
  geom_boxplot() +
  scale_fill_manual(values = diagnosis_colors) +
  labs(title = "Perimeter Mean by Diagnosis", x = "Diagnosis", y = "Perimeter Mean")
```

With median standardized values of 0.92 (malignant) and -0.57 (benign), perimeter_mean provides strong separation across diagnosis groups. This aligns with the biological interpretation that malignant tumors tend to exhibit more irregular and extended boundaries.

### Area Mean by Diagnosis

```{r}
ggplot(Cancer_Data, aes(x = diagnosis, y = area_mean, fill = diagnosis)) +
  geom_boxplot() +
  scale_fill_manual(values = diagnosis_colors) +
  labs(title = "Area Mean by Diagnosis", x = "Diagnosis", y = "Area Mean")
```

Area_mean is another high-separation variable, with medians of 0.79 (malignant) vs. -0.56 (benign). This confirms that overall tumor size plays a major role in classification and supports including this feature in both GLM and Random Forest models.

### Concave Points Worst by Diagnosis

```{r}
ggplot(Cancer_Data, aes(x = diagnosis, y = concave.points_worst, fill = diagnosis)) +
  geom_boxplot() +
  scale_fill_manual(values = diagnosis_colors) +
  labs(title = "Concave Points Worst by Diagnosis", x = "Diagnosis", y = "Concave Points (Worst)")
```

Concave points (worst) provides the strongest contrast of all features (median = 1.03 for malignant vs. -0.61 for benign), indicating more severe contour indentations in malignant tumors. Its discriminative power suggests it could dominate variable importance rankings.

### Tumor Volume by Diagnosis

```{r}
ggplot(Cancer_Data, aes(x = diagnosis, y = tumor_volume, fill = diagnosis)) +
  geom_boxplot() +
  scale_fill_manual(values = diagnosis_colors) +
  labs(title = "Tumor Volume by Diagnosis", x = "Diagnosis", y = "Tumor Volume")
```

The derived tumor_volume feature—based on radius, perimeter, and area—shows a clean class split: 0.61 for malignant vs. -0.54 for benign. This validates our domain-informed feature engineering step and demonstrates that volume integrates multiple shape signals effectively.

### Texture Worst by Diagnosis

```{r}
ggplot(Cancer_Data, aes(x = diagnosis, y = texture_worst, fill = diagnosis)) +
  geom_boxplot() +
  scale_fill_manual(values = diagnosis_colors) +
  labs(title = "Texture Worst by Diagnosis", x = "Diagnosis", y = "Texture Worst")
```

Although less extreme than structural metrics, texture_worst still displays a meaningful shift (0.53 malignant vs. -0.47 benign). Its relevance may lie in capturing the heterogeneity of cell nucleus surfaces, supporting its inclusion in models prioritizing subtle distinctions.

### Smoothness Worst by Diagnosis

```{r}
ggplot(Cancer_Data, aes(x = diagnosis, y = smoothness_worst, fill = diagnosis)) +
  geom_boxplot() +
  scale_fill_manual(values = diagnosis_colors) +
  labs(title = "Smoothness Worst by Diagnosis", x = "Diagnosis", y = "Smoothness Worst")
```

Smoothness worst exhibits the smallest—but still present—class-based median shift (0.49 malignant vs. -0.31 benign). While its predictive power may be secondary, its potential contribution to ensemble or regularized models remains valuable.



## Model Development: Training and Validation Split

To evaluate model generalizability, we partitioned the dataset into an 80% training set and a 20% validation set. The random seed was fixed to ensure reproducibility of results.

```{r}
#Fix the random seed for reproducibility
set.seed(123)  
n <- nrow(Cancer_Data)

#Randomly select 80% of the data to use as training set
index <- sample(1:n, size = 0.8 * n)  

#Create 80/20 training and validation sets
Cancer_Data_train <- Cancer_Data[index, ] 
validation_data <- Cancer_Data[-index, ] 
```


# Logistic Regression Model (GLM)

## Stepwise Logistic Model (All Predictors Included)
```{r, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
#Define predictor set
predictors <- setdiff(names(Cancer_Data_train), c("id", "diagnosis", "diagnosis_bin", "predicted_prob","Hospital", "tumor_volume"))

#Create full and null formulas
full_formula <- as.formula(paste("diagnosis_bin ~", paste(predictors, collapse = " + ")))
full_model <- glm(full_formula, data = Cancer_Data_train, family = binomial)
null_model <- glm(diagnosis_bin ~ 1, data = Cancer_Data_train, family = binomial)

#Stepwise selection using AIC & BIC
step_aic <- step(null_model, 
                 scope = list(lower = null_model, upper = full_model),
                 direction = "both",
                 trace = TRUE)

step_bic <- step(null_model,
                 scope = list(lower = null_model, upper = full_model),
                 direction = "both",
                 k = log(n),
                 trace = TRUE)
```

## Stepwise Logistic Model (Excluding Tumor Volume Predictors)

```{r}
#Define predictor set
predictors <- setdiff(names(Cancer_Data_train), c("id", "diagnosis", "diagnosis_bin", "predicted_prob","Hospital",
  "radius_mean", "perimeter_mean", "area_mean",
  "radius_se", "perimeter_se", "area_se",
  "radius_worst", "perimeter_worst", "area_worst"))

#Create full and null formulas
full_formula_notumor <- as.formula(paste("diagnosis_bin ~", paste(predictors, collapse = " + ")))
full_model_notumorralate <- glm(full_formula_notumor, data = Cancer_Data_train, family = binomial)
null_model_notumorralate <- glm(diagnosis_bin ~ 1, data = Cancer_Data_train, family = binomial)

#Stepwise selection using AIC & BIC
step_aic_notumorralate <- step(null_model_notumorralate, 
                 scope = list(lower = null_model_notumorralate, upper = full_model_notumorralate),
                 direction = "both",
                 trace = TRUE)


step_bic_notumorralate <- step(null_model_notumorralate, 
                 scope = list(lower = null_model_notumorralate, upper = full_model_notumorralate),
                 direction = "both",
                 k = log(n), 
                 trace = TRUE)
```


## Model selection
```{r}
#AIC-selected model 
summary(step_aic)
#BIC-selected model
summary(step_bic)

#AIC summary with tumor volume 
summary(step_aic_notumorralate)
#BIC summary with tumor volume 
summary(step_bic_notumorralate)
```


## Model Comparison via ROC and Confusion Matrix


```{r}
#Model A1: Without tumor_volume (complex)
model_A1 <- glm(formula = diagnosis_bin ~ perimeter_worst + smoothness_worst + 
                             texture_worst + radius_se + compactness_mean + symmetry_worst + 
                             concave.points_mean + texture_se + concavity_worst + compactness_se, 
                family = binomial, data = Cancer_Data_train)

#Model A2: Without tumor_volume (simplified)
model_A2 <- glm(formula = diagnosis_bin ~ perimeter_worst + smoothness_worst + 
                             texture_worst + radius_se + concave.points_worst + compactness_mean, 
                family = binomial, data = Cancer_Data_train)

#Model B1: With tumor_volume (complex)
model_B1 <- glm(formula = diagnosis_bin ~ tumor_volume + texture_worst + 
                             smoothness_worst + symmetry_worst + concave.points_mean + 
                             compactness_mean + fractal_dimension_worst, 
                family = binomial, data = Cancer_Data_train)

#Model B2: With tumor_volume (simplified)
model_B2 <- glm(formula = diagnosis_bin ~ concave.points_worst + tumor_volume + 
                             texture_worst + smoothness_worst, 
                family = binomial, data = Cancer_Data_train)


#Define evaluation function
evaluate_model <- function(model, valid_data, model_name = "Model") {
  # redictive probability
  valid_data$predicted_prob <- predict(model, newdata = valid_data, type = "response")
  
  #Predict classes (threshold = 0.5)
  valid_data$predicted_class <- ifelse(valid_data$predicted_prob >= 0.5 , 1, 0)
  
  #Pompute ROC and AUC
  roc_obj <- roc(valid_data$diagnosis_bin, valid_data$predicted_prob)
  auc_val <- auc(roc_obj)
  
  #Plot ROC curve
  plot(roc_obj, col = "blue", main = paste("ROC -", model_name))
  legend("bottomright", legend = paste("AUC =", round(auc_val, 3)), 
         col = "blue", lwd = 2, bty = "n")
  
  #Confusion Matrix
  valid_data$predicted_class <- factor(valid_data$predicted_class, levels = c(0, 1))
  valid_data$diagnosis_bin <- factor(valid_data$diagnosis_bin, levels = c(0, 1))
  
  conf <- confusionMatrix(valid_data$predicted_class, valid_data$diagnosis_bin)
  print(conf)
  
  return(auc_val)
}
#Evaluate all models
auc_A1 <- evaluate_model(model_A1, validation_data, "Model A1")
auc_A2 <- evaluate_model(model_A2, validation_data, "Model A2")
auc_B1 <- evaluate_model(model_B1, validation_data, "Model B1")
auc_B2 <- evaluate_model(model_B2, validation_data, "Model B2")
```

## Likelihood Ratio Test: B2 vs. Tumor Volume Only

```{r, warning=FALSE}
#Compare nested models (B2 vs. tumor_volume only) using likelihood ratio test
model_small <- glm(diagnosis_bin ~ tumor_volume, data = Cancer_Data_train, family = binomial)
anova(model_small, model_B2, test = "Chisq")
```


Adding `concave.points_worst`, `texture_worst`, and `smoothness_worst` significantly improved the model fit compared to using `tumor_volume` alone. Based on the likelihood ratio test, the full model (B2) provides a better fit and should be preferred.

The final GLM model achieved an AUC of 0.984, indicating excellent performance in predicting diagnosis. Additional metrics such as accuracy, sensitivity, and specificity are available in the confusion matrix output.


$$
\log\left(\frac{\Pr(\text{diagnosis_bin} = 1)}{\Pr(\text{diagnosis_bin} = 0)}\right) 
= \beta_0 
+ \beta_1 \cdot \text{concave.points_worst} 
+ \beta_2 \cdot \text{tumor_volume} 
+ \beta_3 \cdot \text{texture_worst} 
+ \beta_4 \cdot \text{smoothness_worst}
$$


## kNN Classification

```{r, warning=FALSE}
#Define predictor variables
xvars <- c(
  "radius_mean", "texture_mean", "perimeter_mean", "area_mean", "smoothness_mean",
  "compactness_mean", "concavity_mean", "concave.points_mean", "symmetry_mean", "fractal_dimension_mean",
  "radius_se", "texture_se", "perimeter_se", "area_se", "smoothness_se",
  "compactness_se", "concavity_se", "concave.points_se", "symmetry_se", "fractal_dimension_se",
  "radius_worst", "texture_worst", "perimeter_worst", "area_worst", "smoothness_worst",
  "compactness_worst", "concavity_worst", "concave.points_worst", "symmetry_worst", "fractal_dimension_worst","tumor_volume"
)


#KNN tuning: search k from 1 to 100 and record accuracy
k_values <- 1:100
accuracy_values <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  knn_res <- knn(train = Cancer_Data_train[, xvars, drop = FALSE],
                 test = validation_data[, xvars, drop = FALSE],
                 cl = Cancer_Data_train$diagnosis,
                 k = k_values[i])
  
  accuracy_values[i] <- mean(knn_res == validation_data$diagnosis)
}

#Find optimal k with highest validation accuracy
accuracy_df <- data.frame(k = k_values, accuracy = accuracy_values)
best_k <- accuracy_df$k[which.max(accuracy_df$accuracy)]
best_accuracy <- max(accuracy_df$accuracy)

#Plot accuracy vs. k
ggplot(accuracy_df, aes(x = k, y = accuracy)) +
  geom_line(color = "blue") +
  geom_point(aes(x = best_k, y = best_accuracy), color = "red", shape = 17, size = 3) +
  labs(title = "KNN Classification Accuracy by k",
       x = "k (Number of Neighbors)",
       y = "Accuracy") +
  theme_minimal()

#Final prediction using best k
final_knn <- knn(train = Cancer_Data_train[, xvars, drop = FALSE],
                 test = validation_data[, xvars, drop = FALSE],
                 cl = Cancer_Data_train$diagnosis,
                 k = best_k,
                 prob = TRUE)

#Save predictions
validation_data$pred_diagnosis <- final_knn

#Save probability of "M" (malignant)
#If prediction is "B", prob attribute is for "B" → so we flip it
validation_data$prob_M <- ifelse(final_knn == "M", attr(final_knn, "prob"), 1 - attr(final_knn, "prob"))

roc_curve <- roc(response = validation_data$diagnosis,
                 predictor = validation_data$prob_M,
                 levels = c("B", "M"),  # Specify levels for correct direction
                 direction = "<")

auc_value <- auc(roc_curve)

plot(roc_curve, col = "blue", main = "ROC Curve for kNN")
  legend("bottomright", legend = paste("AUC =", round(auc_value, 3)),
         col = "blue", lwd = 2, bty = "n")

#Print best k and accuracy
cat("Best k value:", best_k, "\n")
cat("Validation Accuracy:", round(best_accuracy, 3), "\n")

#Evaluate KNN performance
true_labels <- factor(validation_data$diagnosis, levels = c("B", "M"))
predicted_labels <- factor(validation_data$pred_diagnosis, levels = c("B", "M"))

#Confusion matrix
cm <- confusionMatrix(predicted_labels, true_labels)
print(cm)

#Print best k and accuracy
cat("Best k value:", best_k, "\n")
cat("Validation Accuracy:", round(best_accuracy, 3), "\n")

#Ensure predicted and true labels are factors with the same levels
true_labels <- factor(validation_data$diagnosis, levels = c("B", "M"))
predicted_labels <- factor(final_knn, levels = c("B", "M"))

#Generate confusion matrix
cm <- confusionMatrix(data = predicted_labels, reference = true_labels)
print(cm)
```

## KNN Classification (Simplified Model)

```{r, warning=FALSE}
#Define simplified predictor set for KNN 
xvars_S <- c("tumor_volume","concave.points_worst" , "texture_worst" , "smoothness_worst")

#Loop over a range of k values to evaluate accuracy
k_values <- 1:100
accuracy_values <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  knn_res <- knn(train = Cancer_Data_train[, xvars_S, drop = FALSE],
                 test  = validation_data[, xvars_S, drop = FALSE],
                 cl    = Cancer_Data_train$diagnosis,
                 k     = k_values[i],
                 prob = TRUE)
  
  accuracy_values[i] <- mean(knn_res == validation_data$diagnosis)
}

#Identify the best k value based on maximum accuracy
accuracy_df <- data.frame(k = k_values, accuracy = accuracy_values)
best_k <- accuracy_df$k[which.max(accuracy_df$accuracy)]
best_accuracy <- max(accuracy_df$accuracy)

#Visualize how accuracy changes with k
ggplot(accuracy_df, aes(x = k, y = accuracy)) +
  geom_line(color = "blue") +
  geom_point(aes(x = best_k, y = best_accuracy), color = "red", shape = 17, size = 3) +
  labs(title = "KNN Accuracy using 4 Variables Model",
       x = "k (Number of Neighbors)", y = "Validation Accuracy") +
  theme_minimal()

#Save probability of "M" (malignant)
#If prediction is "B", prob attribute is for "B" → so we flip it
validation_data$prob_M <- ifelse(knn_res == "M", attr(knn_res, "prob"), 1 - attr(knn_res, "prob"))

#Plot ROC and compute AUC
roc_curve <- roc(response = validation_data$diagnosis,
                 predictor = validation_data$prob_M,
                 levels = c("B", "M"),  # Specify levels for correct direction
                 direction = "<")

auc_value <- auc(roc_curve)

plot(roc_curve, col = "blue", main = "ROC Curve for kNN Using 4 Variable Model")
  legend("bottomright", legend = paste("AUC =", round(auc_value, 3)),
         col = "blue", lwd = 2, bty = "n")

cat("Best k using 4 variables:", best_k, "\n")
cat("Validation Accuracy:", round(best_accuracy, 3), "\n")

#Confusion matrix
true_labels <- factor(validation_data$diagnosis, levels = c("B", "M"))
predicted_labels <- factor(knn_res, levels = c("B", "M"))
cm <- confusionMatrix(data = predicted_labels, reference = true_labels)
print(cm)
```

We used the best k value of 5, selected through validation. With an 80/20 training/validation split and a fixed random seed, the final KNN model achieved a validation accuracy of 0.974. Additional performance metrics, including sensitivity and specificity, are reported in the confusion matrix.




## Multilevel Model


### Research Question for Multilevel Modeling

This multilevel modeling part explores the following research question:  

**Does the predictive power of tumor-related features vary across latent population structures in breast cancer diagnosis?**  

If such heterogeneity exists, can we identify these structural groupings through techniques like PCA or K-means clustering, and account for them using mixed-effects modeling?  

Multilevel modeling is introduced here to address inter-group variability that traditional models cannot capture, allowing us to assess whether such structure has statistical explanatory power.




## PCA 

To explore potential structural groupings, we begin with dimensionality reduction via Principal Component Analysis (PCA).

```{r}
#Select relevant variables, excluding ID, diagnosis, and Hospital
vars <- setdiff(names(Cancer_Data_train), c("id", "diagnosis", "diagnosis_bin", "predicted_prob","Hospital",
  "radius_mean", "perimeter_mean", "area_mean",
  "radius_se", "perimeter_se", "area_se",
  "radius_worst", "perimeter_worst", "area_worst"))

#Filter data and remove rows with missing values
cancer_filtered <- Cancer_Data_train %>%
  dplyr::select(all_of(vars)) %>%
  filter(if_all(everything(), ~ !is.na(.))) %>%  
  mutate(across(everything(), as.numeric))      

#Perform PCA on the standardized data
pca_model <- prcomp(cancer_filtered, scale. = TRUE)

#Print PCA loadings
loadings <- pca_model$rotation
print("~~PCA Loadings~~")
print(round(loadings, 3))

#Print proportion of variance explained by each principal component
explained_var <- summary(pca_model)$importance
print("~~Proportion of variance explained by each PC~~")
print(round(explained_var, 3))



#Visualize variable contributions to principal components
fviz_pca_var(pca_model, 
             col.var = "contrib",
             gradient.cols = c("blue", "orange", "red"),
             repel = TRUE) +
  ggtitle("PCA Loadings - Cancer Data (Excluding Hospital)")


#Visualize samples colored by cos2
fviz_screeplot(pca_model, addlabels = TRUE, ylim = c(0, 50))


fviz_pca_ind(pca_model, 
             geom.ind = "point", 
             col.ind = "cos2",
             gradient.cols = c("blue", "orange", "red"),
             repel = TRUE) +
  ggtitle("PCA of Samples (Individuals) - Cancer Data")



#Extract PC1 and PC2 scores
pca_scores <- as.data.frame(pca_model$x[, 1:2])  
colnames(pca_scores) <- c("PC1", "PC2")

#Apply k-means clustering (2 clusters)
set.seed(123)  
kmeans_result <- kmeans(pca_scores, centers = 2, nstart = 25)

#Add cluster labels to PCA scores
pca_scores$Cluster <- as.factor(kmeans_result$cluster)

#Visualize clusters on PC1-PC2 plot
ggplot(pca_scores, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(title = "K-means Clustering on PCA Results",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  scale_color_manual(values = c("red", "blue"))  



#Extract loading matrix from PCA result
loadings_matrix <- pca_model$rotation

#Initialize an empty list to store top contributing variables for each PC
top_variables_list <- list()

#Get top 10 contributing variables for each of the first 7 PCs
for (i in 1:7) {
  pc_loadings <- loadings_matrix[, i]
  top_vars <- sort(abs(pc_loadings), decreasing = TRUE)[1:10]
  top_variables_list[[paste0("PC", i)]] <- names(top_vars)
}

#Display the top variables for each of the first 7 PCs
top_variables_list
```


## PCA Grouping Based on PC1 and PC2 Scores

```{r}
#Add cluster group from k-means result
pca_scores$cluster_group <- as.factor(kmeans_result$cluster)

#Group PC1 into tertiles: Low, Medium, High
pca_scores$pc1_group <- cut(pca_scores$PC1,
                            breaks = quantile(pca_scores$PC1, probs = c(0, 0.33, 0.66, 1)),
                            labels = c("Low", "Medium", "High"),
                            include.lowest = TRUE)


#Bin PC1 and PC2 separately into two levels: Low and High
pca_scores$pc1_bin <- cut(pca_scores$PC1, breaks=2, labels=c("Low", "High"))
pca_scores$pc2_bin <- cut(pca_scores$PC2, breaks=2, labels=c("Low", "High"))

#Combine PC1 and PC2 binary groups 
pca_scores$pca_group <- interaction(pca_scores$pc1_bin, pca_scores$pc2_bin)

#Add PCA group and cluster group as grouping variables
pca_scores$diagnosis_bin <- Cancer_Data_train$diagnosis_bin
pca_scores$tumor_volume <- Cancer_Data_train$tumor_volume
pca_scores$concave.points_worst <- Cancer_Data_train$concave.points_worst
pca_scores$texture_worst <- Cancer_Data_train$texture_worst
pca_scores$smoothness_worst <- Cancer_Data_train$smoothness_worst

#Variables used in modeling
tumor_shape_vars <- c("concave.points_mean", "concave.points_worst", 
                      "concavity_mean", "concavity_worst")


#Multilevel model with PCA group as random intercept
model1 <- glmer(diagnosis_bin ~ tumor_volume + (1 | pca_group), 
               data = pca_scores, 
               family = binomial)

#Multilevel model with cluster group as random intercept
model2 <- glmer(diagnosis_bin ~ tumor_volume + (1 | cluster_group), 
               data = pca_scores, 
               family = binomial)

#View summaries and random effects
summary(model1)
summary(model2)
ranef(model1)
ranef(model2)
```

The effect of `tumor_volume` on malignant diagnoses was highly significant in both models. However, the clustering-based model demonstrated better overall fit (lower AIC and BIC), suggesting that naturally derived structural groupings explain more variation in diagnostic outcomes than manually defined regional bins.

Fixed effects represent overall effects that apply to all observations. Here, the effect of tumor_volume is consistently strong and statistically significant.

Random effects capture variability across latent groupings that are not the main focus but influence the response. These are included to control for unobserved heterogeneity.

Among the PCA-defined groups:

Low.High (small volume, irregular shape/high texture) shows the highest baseline risk,

while High.Low (large volume, smooth shape/low texture) shows the lowest.


Grouping Explanation of possible combinations of features

Low.Low Small volume + Smoothness/weak texture

High.Low large volume + smooth/weak texture

Low.High Small volume + irregular edges/high texture

High.High large volume + irregular edges/high texture


## LASSO Regression

```{r}
#Prepare data for glmnet
y <- Cancer_Data$diagnosis
X <- as.matrix(Cancer_Data[, xvars_S])

#Fit LASSO logistic regression with cross-validation
set.seed(123)
lasso_fit <- cv.glmnet(X, y, family = "binomial", alpha = 1)

#Plot cross-validation error to find optimal lambda
plot(lasso_fit)

#Extract lambda that minimizes CV error
best_lambda <- lasso_fit$lambda.min
cat("Best lambda :", best_lambda)

#Get coefficients at optimal lambda
coef(lasso_fit, s = "lambda.min")
```

The optimal lambda is very small, indicating minimal regularization. This suggests the model behaves almost like a standard logistic regression, so LASSO adds little benefit in this case.

## Random Forest

```{r}
#Convert diagnosis to factor for classification
Cancer_Data$diagnosis <- as.factor(Cancer_Data$diagnosis)

#Train a Random Forest model with 500 trees and variable importance enabled
set.seed(123)
rf_model <- randomForest(diagnosis ~ radius_mean + texture_mean + perimeter_mean + area_mean + smoothness_mean,
                         data = Cancer_Data,
                         importance = TRUE,
                         ntree = 500)

#View model summary
print(rf_model)

#Variable importance plot
varImpPlot(rf_model)
```

Here, we used a random forest model and obtained an out-of-bag (OOB) error rate of 7.91%, indicating that the model predicts the correct diagnosis approximately 92.1% of the time. Based on the confusion matrix, the model performs better at identifying benign tumors than malignant ones, possibly because benign cases are more prevalent in the dataset.

In the variable importance plot, `texture_mean` shows the highest value, suggesting that prediction accuracy significantly decreases when this feature is randomly permuted. This indicates that `texture_mean` is one of the most influential predictors. In contrast, `radius_mean` has the lowest importance score, implying that the model relies less on this variable when making predictions.



